RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-08 formalframework.tex
added small note formal framework
improved formal framework

In Section~\ref{sect:gmas} we developed a list of critical questions and argument schemes by analyzing transcripts of discussions about the development of a traffic simulator. The resulting list is shown in Table~\ref{table:argument-schemes}. We also discussed various examples of applications of the different critical questions and all four different effects (right column of Table~\ref{table:argument-schemes}): \textsf{INTRO}, \textsf{DISABLE}, \textsf{REPLACE}, and \textsf{ATTACK}.

The examples and corresponding visualizations of the previous sections provide some insight on how our RationalGRL framework can be used. In this section, we present a more precise formalization of our framework. In the first two subsections we formalize a static representation of our framework: We develop a formal language to specify a GRL model (Section \ref{sect:goalmodeling:logicallanguage}) and we formalize the notion of an argument (Section \ref{sect:goalmodeling:argumentationsemantics}). In the third subsection we formalize the dynamics of our framework: we develop algorithms for instantiating argument schemes and for answering critical questions. The formal framework we develop in this section serves as the underlying model for the prototype implementation we present in the next section.

\subsection{Logical Language for RationalGRL}
\label{sect:goalmodeling:logicallanguage}

We assume familiarity with basic notions from propositional logic and set theory. An \emph{atom} is a formula with no logical connectives ($\vee, \wedge, \rightarrow, \leftrightarrow$) or negation ($\neg$). We first define the logical language for RationalGRL.

\begin{definition}[RationalGRL Language]
The \emph{RationalGRL language} $\mathcal{L}$ contains the following atoms:
\begin{itemize}
\item $actor(i)$: Identifier $i$ is an actor.
\item $softgoal(i)$: Identifier $i$ is a softgoal.
\item $goal(i)$: Identifier $i$ is a goal.
\item $task(i)$: Identifier $i$ is a task.
\item $resource(i)$: Identifier $i$ is a resource.
\item $decomp(k,i,J,dtype)$: Identifier $k$ is a $dtype$-decomposition (and/or/xor) from the element corresponding to identifier $i$ to the set of elements corresponding to identifiers $J$.
\item $dep(k,i,j)$: Identifier $k$ is a dependency link from the element corresponding to identifier $i$ to the element corresponding to identifier $j$.
\item $contr(k,i,j,ctype)$: Identifier $k$ is a $ctype$-contribution (+/-) from the element corresponding to identifier $i$ to the element corresponding to identifier $j$.
\item $has(i,j)$: Identifier $i$ (which is an actor) has the element corresponding to identifier $j$.
\item $disabled(i)$: The element or relationships corresponding to identifier $i$ is disabled.
\item $name(i,p)$: The name of the element corresponding to identifier $i$ is $p$.
\end{itemize}

We use integers $i,j,k\in\mathbb{N}$ as identifiers for actors and GRL elements. For all of the above definitions we assume that all identifiers used in each atom are different, i.e.,
\begin{itemize}
\item $k\not=i$,$k\not=i$,$k\not=j$
\item forall $j\in J: k\not=j, i\not=j$
\end{itemize}
\end{definition}

Using the language from Definition 1, we can define GRL models. A \emph{ground atom} is an atom without variables.

\begin{definition}[GRL model] \label{def:argument}
A \emph{GRL model} $\mathcal{M}$ is a set of ground atoms based on $\mathcal{L}$.
\end{definition}

An example of the specification of the GRL model in Figure~\ref{fig:examples:decomposition} is shown in Table~\ref{table:grl_atom_spec}, written in logic programming style. In this article we do not make use of any logic programming techniques but this would be interesting future work (see Section~\ref{sect:goalmodeling:openissues}). A more elaborate example of a specification is shown in Appendix~\ref{ch:grlspec}, showing a complete specification of the traffic simulator GRL model in Figure~\ref{fig:trafficsim}.

\begin{table}[h!]
\centering
\begin{tabularx}{0.4\textwidth}{|X|}
\hline
\texttt{goal(0).}\\
\texttt{task(1).}\\
\texttt{task(2).}\\
\texttt{name(0,simulate).}\\
\texttt{name(1,static\_simulation).}\\
\texttt{name(2,dynamic\_simulation).}\\
\texttt{decomp(3,0,[1,2],or).}\\
\hline
\end{tabularx}
\caption{Specification of the GRL model in Figure~\ref{fig:examples:decomposition}}
\label{table:grl_atom_spec}
\end{table}

\subsection{Formal argumentation}
\label{sect:goalmodeling:argumentationsemantics}

In the previous subsection we introduced an atomic language to specify a GRL model. In this subsection we give a formal definition of an argument, which is simply a set of atoms from our language.

\begin{definition}[Argument] \label{def:argument}
An \emph{argument} $A\subseteq Args$ is a set of ground atoms based on $\mathcal{L}$.
\end{definition}

This simple definition allows us to form arguments about (parts of) a GRL model. For instance, $$\{goal(0),name(0,development\_team)\},$$ is an argument stating that identifier 0 is a goal element with name ``development team''. Notice that an argument and a GRL model are essentially the same: as was discussed in Section \ref{sect:metamodel}, each element of a GRL model is an argument.

Definition \ref{def:argument} does not put any restrictions on the content of an argument. This means it is possible to form arguments that do not correspond to correct or sensible GRL models. Consider for instance the following argument: $$\{goal(0), task(0), name(0,static\_simulation)\}.$$ This argument states that identifier 0 is both a goal and a task with name ``static simulation'', which is clearly undesirable since it does not correspond to a valid GRL model. While we could have chosen to put logical constraints on the content of an argument, we choose not to do so here. Instead, we put these restrictions on the algorithms in section~\ref{sect:algorithms}. The algorithms are constructed in such a way that they only allow the construction of arguments that correspond to valid GRL models. The advantage of this approach is that it keeps the presentation of the formal part relatively simple.

In line with the main approaches in formal argumentation, we next introduce an argumentation framework as a set of arguments and an attack relation between the arguments.

\begin{definition}[Argumentation framework~\cite{Dung1995}] \label{def:argumentationframework}
An argumentation framework $AF=(Args,Att)$ consists of a set of arguments $Args$ and an attack relationship $Att:Args\times Args$, where $(A_1,A_2)\in Att$ means that argument $A_1\in Args$ attacks arguments $A_2\in Args$.
\end{definition}

@@

\subsection{Algorithms for argument schemes and critical questions}
\label[sect:algorithms}

Let us briefly take stock:
\begin{itemize}
\item We developed an initial set of argument schemes and critical questions in Section~\ref{sect:gmas}, by annotating transcripts.
\item We developed a visual notation to illustrate some argument schemes and critical questions in Section~\ref{sect:examples}.
\item We proposed a logical language consisting of atomic sentence to describe a GRL model in Section~\ref{sect:goalmodeling:logicallanguage}.
\item We formalized an ``argument'' as a set of atoms from our language, and we introduced \emph{argumentation semantics} to compute sets of accepted arguments in Section~\ref{sect:goalmodeling:argumentationsemantics}.
\end{itemize}

In this subsection we develop algorithms for applying argument schemes and critical questions. The process of applying these algorithms and generating goal models from them is shown in context of the RationalGRL methodology in Figure~\ref{fig:methodologyalgorithm} \todo{Marc}{Marc}{Create an extended version of the methodology, containing where which parts of the formal framework are used.} These algorithms produce arguments (Definition~\ref{def:argument}) and attack relations, forming an argumentation framework (Definition~\ref{def:argumentationframework}). One can then use argumentation semantics (Definition~\ref{def:semantics}) to compute sets of accepted arguments. The content of these arguments is then used to compute the resulting RationalGRL model, together with enabled and disabled GRL elements and their underlying arguments.

In the following algorithms, we assume the following global variables:

\begin{definition}[Global variables] The following variables are intended to have a \emph{global} scope and have a \emph{static} value, meaning the lifetime of the variable extends across the entire run of the program.
\begin{itemize}
\item $id$: the current highest identifier of the elements. This variable is increased for each new element that is added.
\item $Args$: contains the set of all arguments.
\item $Att$: contains the set of all attack relations.
\end{itemize}
\end{definition}

\subsubsection*{Algorithms for argument schemes}
The algorithms for the argument schemes consist of forming a new argument and adding it to the set of arguments. No attack relations are introduced.

\begin{algorithm}[h]
  \caption{Applying AS0: Actor $a$ is relevant}\label{alg:as0}
  \begin{algorithmic}[1]
    \Procedure{$AS_0$}{$a$}
    \State $id\gets id+1$\label{alg:as0:1}
    \State $A \gets \{actor(id), name(id,a)\}$\label{alg:as0:2}
    \State $Args \gets Args \cup A$\label{alg:as0:3}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\emph{Algortihm~\ref{alg:as0} for argument scheme AS0}: The algorithm takes one argument, namely the name of the actor $a$. On line~\ref{alg:as0:1} of the algorithm, the global variable $id$ is increased by one. This ensures that each new argument has a unique identifier. On line~\ref{alg:as0:2}, the argument for the actor is formed, consisting of two statements, stating respectively that $id$ is an actor, and that the name of this actor is $a$. Finally, on line~\ref{alg:as0:3} this argument is added to the global set of arguments $Args$. In Figure~\ref{fig:examples:relevant-actor2}, the application of the argument scheme $AS_0(\text{Development team})$, results in one argument: $$Args = \{\{actor(0), name(0, \text{Development team})\}\}.$$

@@

\emph{Algorithm~\ref{alg:as1} for argument scheme AS1:} This argument scheme takes two arguments, the identifier $a_{id}$ of the actor and the resource name $n$. The algorithm is similar to the previous one, with the difference that the newly added argument contains the statement $has(a_{id},id)$ as well, meaning that the actor with id $a_{id}$ has element $id$ (which is a resource). As an example, let us formalize the argument corresponding to resource \emph{External library} of actor \emph{Traffic tycoon} in Figure~\ref{fig:transcripts:grl}. First, we assume some id is associated with the actor: $$\{actor(0),name(0,traffic\_tycoon)\}.$$ Then we can formalize the argument for the resource as follows: $$\{resource(1),name(1,external\_library),has(0,1)\}.$$

@@

\emph{Algorithm~\ref{alg:as5} for argument scheme AS5:} The procedure in Algorithm~\ref{alg:as5} takes three arguments: $g_{id}$ is the identifier of goal $G$, $T=(T_1,\ldots,T_n)$ is a list of decomposing task names, and $type\in\{and,or,xor\}$ is the decomposition type. The difficulty of this algorithm is that each of the tasks are stated in natural language, and it is not directly clear whether these tasks are already in the GRL model or not. Therefore, we have to check for each tasks whether it already exists, and if not, we have to create a new task. On line~\ref{alg:as5:1}, the set $T_{id}$ is initialized, which will contain the ids of the tasks $T_1,\ldots,T_n$ to decompose into. In the for-loop, the if-statement on line~\ref{alg:as5:2} checks whether some argument already exists for the current task $T_i$, and if so, it simply adds the identifier of the task ($t_{id}$) to the set of task identifiers $T_{id}$.\footnote{Note that the existing task may be disabled, but this does not matter, since the decomposition relation will be suppressed for disabled elements.} Otherwise (line~\ref{alg:as5:3}), a new task is created and the new identifier $id$ is added to the set of task identifiers. After the for loop on line~\ref{alg:as5:4}, an argument for the decomposition link itself is constructed, and it is added to the set of arguments $Args$.

Let us explain this algorithm with the XOR-decomposition of goal \texttt{Generate cars} of Figure~\ref{fig:transcripts:grl}. Suppose the following arguments are constructed already:
\begin{itemize}
\item $\{goal(0),name(0,generate\_cars)\},$
\item $\{taks(1),name(1,keep\_same\_cars\}.$
\end{itemize}
Suppose furthermore that someone wants to put forward the argument that goal \texttt{Generate cars} XOR-decomposes into tasks \texttt{Keep same cars} and \texttt{Create news cars}. Formally: $AS_5(0,\{generate\_cars,keep\_same\_cars\},xor)$. The algorithm will first set $T_{id}=\emptyset$, and then iterate over the two task names. For the first task $generate\_cars$, there does not exist an argument $\{task(t_{id}),name(t_{id},generate\_cars)\}$ yet, so a  new argument is created. Suppose the following argument is created: $\{task(2),name(2,generate\_cars)\}.$ After this, 2 is added to $T_{id}$. For the second task an argument exists already, namely $\{task(1),name(1,keep\_same\_cars)\}$, so 1 is simply added to $T_{id}$. After the for loop, we have $T_{id}=\{1,2\}$. Next, the decomposition argument is created, which is $\{decomp(3,0,\{1,2\},xor)\}$. This argument is added to $Args$ and the algorithm terminates.

@@

\emph{Algorithm~\ref{alg:as6} for argument scheme AS6:} The procedure in Algorithm~\ref{alg:as6} takes two arguments: $t_{id}$ is the identifier of task $T$, and $s$ is the softgoal name that is contributed to. The idea behind this algorithm is very similar to the previous one, with the difference that in the current algorithm we create a single relation, while we created a set of relations in the previous algorithm. First, the if-statement on line~\ref{alg:as6:if} checks whether the softgoal exists already, and if not, an argument is added for it. This ensures that all softgoals have corresponding arguments. After the if-statement, the argument for the contribution link is created and it is added to the set of arguments $Args$.

Let us again illustrate this with a simple example from Figure~\ref{fig:transcripts:grl}. Suppose the following argument exists already: $\{task(0),name(0,keep\_same\_cars\}$, and suppose someone would like to add an argument that the task \texttt{Keep same cars} contributes positively to softgoal \texttt{Dynamic simulation}, i.e. $AS_6(0,dynamic\_simulation)$. The algorithm first checks whether an argument already exists for the softgoal, and when it finds out it does not exist, creates the argument $\{softgoal(1),name(1,dynamic\_simulation)\}$ and adds it to $Args$. Then, the argument for the contribution is added to $Args$ as well, which is $\{contr(2,0,1,pos)\}$.

@@

@@

\subsection{Constructing GRL models}

\todo{F}{F,M}{It would be good to have a formal definition of how to construct models given an argumentation framework. If we define a model M at the start of this section, this should be fairly easy.}

>>>>

In Section~\ref{sect:gmas} we developed a list of critical questions and argument schemes by analyzing transcripts of discussions about the development of a traffic simulator. The resulting list is shown in Table~\ref{table:argument-schemes}. We then presented the RationalGRL framework in secton~\ref{sect:overview}, consisting of a modeling language, a metamodel, and various examples of using the argument schemes and critical questions. In section~\ref{sect:methodology} we explained how this framework can be used by practitioners by explaining the methodology and tool support.

In this section we present a formalization of RationalGRL based on formal logic. This is done for multiple reasons: (i) Most approaches in formal argumentation use formal logic, allowing us to employ existing technique directly in order to compute which arguments are accepted and which are rejected, (ii) we can be more precise about how critical questions are answered, (iii) we can show that RationalGRL models can be translated in to valid GRL models in a very precise way, and (iv) the formal approach is a basis for automating the framework in terms of tool support.

In the first two subsections we formalize a static representation of our framework: We provide a formal specification of a GRL model based on the GRL metamodel (Section \ref{sect:goalmodeling:logicallanguage}) in the first subsection, and we extend this with arguments and attack links in the second subsection, hereby obtaining a formal specification of a RationalGRL model. In the third subsection we turn to the dynamics of our framework. We develop algorithms for instantiating argument schemes and for answering critical questions. We show that each algorithm is correct, meaning that if the RationalGRL model was valid before applying the algorithm, it is also valid afterwards.

\subsection{Formal Specification of GRL}
\label{sect:goalmodeling:logicallanguage}

In this subsection we formalize a GRL model based on the GRL metamodel and the jUCMNav implementation. We first formalize elements of GRL (intentional elements and actors), and then formalize the links.

\subsubsection{Intentional Element and Actors}

We start with the definition of various sets of elements that we use in subsequent definitions.

\begin{definition}[Set definitions]
\label{def:set-definitions}
Throughout this section, we adopt the convention that variables start with a lowercase letter, and sets and constants start with an uppercase.
\label{def:ie}
We define the following sets:
\begin{itemize}
\item $Type = \{Softgoal, Goal, Task, Resource\}$,
\item $Names$ is a finite set of string.
\item $DecompType = \{AND,OR,XOR\}$.
\item $ContribValue = \{Break, Hurt, Some\ nega-$ $tive, Unknown, Some\  positive, Help, Make\},$
\end{itemize}
\end{definition}

Next we define an intentional element.

\begin{definition}[Intentional Element]
\label{def:ie}
An intentional element $ie\in \mathbb{N}\times Type\times Names\times DecompType$ is a relation, where $ie = (id, type, name, decomptype)$ means:
\begin{itemize}
\item $id\in \mathbb{N}$ is a unique identifier for the element,
\item $type\in Type$ specifies the type of the element,
\item $name \in Names$ is a string description of the element,
\item $decomptype\in DecompType$ refers to the type of decomposition.\footnote{Note that the decomposition type is relevant only if the element is in fact decomposed into other elements, but since the jUCMNav implementation defines the decomposition type on the element, we do the same here.}
\end{itemize}
A set of intentional elements is denoted by $IE$.
\end{definition}

The definition above is sufficient to capture all intentional elements used in GRL. However, we present some syntactic sugar in the next definition by abbreviating the definition above in various ways. This does not add anything new to the previous definition, but it simplifies some of the notation.

\begin{definition}[Notation]
\label{def:notation}
We adopt three conventions simplifying our notation:
\begin{itemize}
\item
We refer to the element of an intention element $ie$ using the dot (".") notation. That is, we may refer to the id, type, name and decomposition type with respectively $ie.id$, $ie.type$, $ie.name$, and $ie.decomptype$.
\item 
We refer to an intentional element with $ie.id=i$ simply by $ie_i$.\footnote{Note that the reference $ie_i$ is never ambiguous in GRL since identifiers for intentional elements are always unique (see Definition~\ref{def:grlmodel}).} For instance, we may refer to the intentional element $ie = (0, Goal, \text{Make profit}, AND)$ with $ie_0$ and write $ie_0.type = Goal$, $ie_0.name = \text{Make profit}$, and $ie_0.decompositiontype = AND$. 
\item
We can also refer to intention elements of a specific type simply by $type_{id}$. For instance, we can abbreviate the element in the previous item with $goal_0$ and write $goal_0.name = \text{Make profit}$, and $goal_0.decomptype = AND$.
\end{itemize}
\end{definition}

\begin{definition}[Actor]
\label{def:actor}
An actor $act \in \mathbb{N}\times Names$ is a relation where $act=(id,name)$ means:
\begin{itemize}
\item $id\in\mathbb{N}$ refers to the unique identifier of the actor, 
\item $name$ refers to a string description of its name.
\end{itemize}

Similar to intentional elements, we may refer to $act = (id,name)$ with $act_{id}$ and write $act_{id}.name$ to refer to its name.

A set of actors is denoted by $Act$.
\end{definition}

In the next definition we specify relations between actors and intentional elements. Each actor can own one or multiple intentional elements. We chose to define this simply as a actor-element relation, which is in line with the jUCMNav implementation.

\begin{definition}[Actor-IE Relations]
\label{def:act-ie-relation}
An \emph{Actor-IE relation} $r_{ActIE}\in\mathbb{N}\times\mathbb{N}$ is a relation $(i,j)$ meaning that an actor with id $i$ has intentional element with id $j$.

A set of Actor-IE relations is denoted by $R_{ActIE}$.
\end{definition}

\subsubsection{Links}

At this point we have defined all intentional elements in GRL and a containment relation between actors and intentional elements. We now turn to the GRL links.

\begin{definition}[Contribution Link]
\label{def:link:contrib}
A \emph{contribution} link: $contrib\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}\times Contribvalue$ is a relation such that $contrib = (i,j,k,value)$ means
\begin{itemize}
\item $i$ is the unique identifier of the link,
\item $j$ is the unique identifier of the IE from which the contribution originates,
\item $k$ is the unique identifier of the IE to which is contributed.
\end{itemize}
Intuitively, $contrib = (i,j,k,value)$ means that $ie_i$ contributes to $ie_j$ with $value$.

A set of contribution links is denoted by $Contrib$.
\end{definition}

\begin{definition}[Decomposition Link]
\label{def:link:decomp}
A \emph{decomposition} link: $decomp\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}$ is a relation such that $decomp = (i,j,k)$ means
\begin{itemize}
\item $i$ is the unique identifier of the link,
\item $j$ is the unique identifier of the decomposing IE,
\item $k$ is the unique identifier of the IE that is being decomposed.
\end{itemize}
Intuitively, $decomp = (i,j,k)$ means that $ie_i$ decomposes $ie_j$.\footnote{Note that the decomposition type is defined on $IE_{k}$, see Definition~\ref{def:ie}.}

A set of decomposition links is denoted by $Decomp$.
\end{definition}

\begin{definition}[Dependency Link]
\label{def:link:dependency}
A \emph{dependency} link: $dep\in \mathbb{N}\times\mathbb{N}\times\mathbb{N}$ is a relation such that $dep = (i,j,k)$ means
\begin{itemize}
\item $i$ is the unique identifier of the link,
\item $j$ is the unique identifier of the dependee IE,
\item $k$ is the unique identifier of the dependent IE.
\end{itemize}
Intuitively, $dep = (i,j,k)$ means that $ie_j$ depends on $ie_j$. \todo{Marc}{Sepideh}{Is this correct? For a dependency relation $A\rightarrow B$, who is dependent on who?}

A set of dependency links is denoted by $Dep$.
\end{definition}

Similar to intentional elements and actors, we may refer to links with an id using the subscript notation, e.g., $contrib_0$ is a contribution link with $contrib.id = 0$.

The definitions above come with various implicit assumptions in order to form a valid GRL model. For instance, if an Actor-IE relation $(i,j)$ occurs in a GRL model, then there should exist an actor with id $i$, and there also should exist an IE with id $j$. We make these assumptions explicit in the following definition.

\begin{definition}[GRL Model]
\label{def:grl-model}
A \emph{GRL model} $GRL=(IE, Act, R_{ActIE}, Contr, Decomp, Dep)$ consists of:
\begin{itemize}
\item A set $IE$ of intentional elements (Def.~\ref{def:ie}),
\item A set $Act$ of actors (Def.~\ref{def:actor}),
\item A set $R_{ActIE}$ of Actor-IE relations (Def.~\ref{def:act-ie-relation}),
\item A set $Contr$ of contribution links (Def.~\ref{def:link:contrib}),
\item A set $Decomp$ of decomposition links (Def.~\ref{def:link:decomp}),
\item A set $Dep$ of dependency links (Def.~\ref{def:link:dependency}).
\end{itemize}

Furthermore, the following conditions are satisfied:
\begin{enumerate}
\item ids are globally unique across IEs, Links, and Actors, i.e., let $X,Y\in \{IE,Act, Contr, Decomp, Dep\}$. For all $X_i$ and $Y_j$: if $i=j$ then $X=Y$ and $X_i=Y_j$.
\item All intentional elements of actors exist: $\forall (i,j)\in R_{ActIE}: act_i \in Act \wedge ie_j \in IE$.
\item An intentional element belongs at most to one actor: $\forall ie_i\in IE: |\{(k,i)\in R_{ActIE}\}| \le 1$.
\item Contribution links connect intentional elements: $\forall (i,j,k,value)\in Contrib: \{ie_j,ie_k\}\subseteq IE$.
\item Decomposition links connect intentional elements: $\forall (i,j,k)\in Decomp: \{ie_j,ie_k\}\subseteq IE$.
\item Dependency links connect intentional elements: $\forall (i,j,)\in Dep: \{ie_j,ie_k\}\subseteq IE$.
\end{enumerate}
\end{definition}

\todo{Marc}{Marc}{Provide explanation and example here.}

\subsection{Formal specification of RationalGRL}

In order to develop a logical framework for RationalGRL, we extend the GRL logical framework of the previous section by adding three elements (see Figure~\ref{fig:rationalgrllegend}):
\begin{itemize}
\item A new element called \emph{generic argument},
\item A new link called \emph{attack link},
\item An labeling of the elements, assigning to each element (an IE, link, or generic argument) the label IN (accepted) or OUT (reject).
\end{itemize}

In the next section we explain how the labeling is computed, but in this section we only provide a formal definition.

We start with the new element which we call the \emph{Generic Argument}.

\begin{definition}[Generic Argument]
\label{def:generic-argument}
A generic argument $arg \in \mathbb{N}\times Names$ is a relation such that $arg=(id, name)$ is an argument with unique identifier $id$ and name $name$. We may refer to the argument with id $i$ simply with $arg_i$. A set of arguments is denoted by $Arg$.
\end{definition}

Note that one of the constraints of a GRL model (Def.~\ref{def:grl-model}) is that GRL links (Def.~\ref{def:link:contrib},~\ref{def:link:decomp}, and~\ref{def:link:dependency}) should connect IEs, which means that in GRL generic arguments cannot be connected with GRL links. This is correct, since generic arguments are not part of GRL, so they should also not occur in links.

We next define the attack link, which is the only link that RationalGRL adds to GRL.

\begin{definition}[Attack Link]
\label{def:link:attack}
An attack link $att\in \mathbb{N}\times \mathbb{N}$ is a relation such that $att=(i,j)$ means:
\begin{itemize}
\item $i$ is the unique identifier of the element that is performing that attack,
\item $j$ is the unique identifier of the element that is being attacked.
\end{itemize}
Intuitively, $att=(i,j)$ means that element $i$ is attacking element $j$.
\end{definition}

Note that the previous definition did not put any constraint on which elements can be connected with attack links. This is what we make explicit in the definition of a RationalGRL model later.

We next add the last part of the RationalGRL addition, namely the labeling of elements. This labeling determines whether an element is accepted (which means it shows up in the GRL model if it is an IE or a link), or rejected (which means it won't be shown).

\begin{definition}[RationalGRL labeling]
\label{def:labeling}
A RationalGRL labeling function $Lab:\mathcal{N}\rightarrow {in,out}$ that is total on the ids of the RationalGRL elements. This means that $Lab(id)$ is defined for each $ie.id, contr.id, decomp.id, dep.id, arg.id$.
\end{definition}

\begin{definition}[RationalGRL Model]
\label{def:rationalgrl-model}
A \emph{RationalGRL model} $RationalGRL=(IE, Act, R_{ActIE}, Contr, Decomp, Dep, Arg, Att, Label)$ consists of:
\begin{itemize}
\item A GRL model $(IE, Act, R_{ActIE}, Contr, Decomp,$ $Dep)$
\item A set of generic arguments $Arg$ (Def.~\ref{def:generic-argument}),
\item A set of attack links (Def.~\ref{def:link:attack}).
\item A labeling function $Lab$ (Def.~\ref{def:labeling}).
\end{itemize}

Furthermore, the following conditions are satisfied. In the following, let $Elems = \{IE, Act, Contr, Decomp, Dep, Arg\}$:
\begin{enumerate}
\item All conditions on a GRL model are satisfied, (Def.~\ref{def:grl-model}),
\item ids are globally unique across IEs, Links, Actors, and Generic arguments, i.e., let $X,Y\in Elems$. For all $X_i$ and $Y_j$: if $i=j$ then $X=Y$ and $X_i=Y_j$.
\item Attack links connect generic arguments, IEs, actors, and links with each other, i.e., let   $\forall (i,j)\in Att: \exists X\in Elems$ s.t. $X_i\in X$ and $\exists Y\in Elems$ s.t. $Y_i\in Y$.
\end{enumerate}
\end{definition}

\todo{Marc}{Marc}{Explain conditions here and give some examples}

\subsubsection{Argumentation semantics}

@@

\todo{Marc}{Marc}{Give example with concrete RationalGRL model}

\subsection{Algorithms for argument schemes}
\label{sect:algorithms}

In the previous section we formalize a \emph{static} representation of the RationalGRL framework. In this section we formalize the \emph{dynamics}. We do so by developing algorithms for applying argument schemes and critical questions in the context of a RationalGRL model~\ref{def:rationalgrl-model}. These algorithms produce new elements and attack relations. We can then use argumentation semantics (Definition~\ref{def:semantics}) to compute sets of accepted arguments. The content of these arguments is then used to compute the resulting RationalGRL model, together with enabled and disabled GRL elements and their underlying arguments.

The algorithms for the argument schemes consist of forming a new argument and adding it to the set of arguments. No attack relations are introduced.

\begin{algorithm}[h]
  \caption{Applying AS0: Actor $a$ is relevant}\label{alg:as0}
  \begin{algorithmic}[1]
    \State $RationalGRL=(IE, Act, R_{ActIE}, Contr, Decomp,$
    \State \qquad\qquad\qquad\qquad$Dep, Arg, Att, Label)$
    \Procedure{$AS_0$}{$a$}
    \State $id\gets mintId()$\label{alg:as0:1}
    \State $Act \gets Act \cup (id, a)$\label{alg:as0:2}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\subsubsection{Algortihm~\ref{alg:as0} for argument scheme AS0}

\paragraph{Explanation} For all of the argument schemes and critical questions algorithms, we assume an underlying RationalGRL model as defined in Def.~\ref{def:rationalgrl-model}. This model is shown in the first two lines.

The algorithm takes one argument, namely the name of the actor $a$. On line~\ref{alg:as0:1} of the algorithm, a new (unique) id is minted. On line~\ref{alg:as0:2}, the the tuple $(id,a)$ is added to the set of actors $Act$, meaning that a new actor with id $id$ and name $a$ is added to the RationalGRL model (Def.~\ref{def:actor}). In Figure~\ref{fig:examples:relevant-actor2}, the application of the argument scheme $AS_0(\text{Development team})$, results in one argument: $$Args = \{\{actor(0), name(0, \text{Development team})\}\}.$$

@@

\paragraph{Example} \emph{Algorithm~\ref{alg:as1} for argument scheme AS1:} This argument scheme takes two arguments, the identifier $a_{id}$ of the actor and the resource name $n$. The algorithm is similar to the previous one, with the difference that the newly added argument contains the statement $has(a_{id},id)$ as well, meaning that the actor with id $a_{id}$ has element $id$ (which is a resource). As an example, let us formalize the argument corresponding to resource \emph{External library} of actor \emph{Traffic tycoon} in Figure~\ref{fig:transcripts:grl}. First, we assume some id is associated with the actor: $$\{actor(0),name(0,traffic\_tycoon)\}.$$ Then we can formalize the argument for the resource as follows: $$\{resource(1),name(1,external\_library),has(0,1)\}.$$

\paragraph{Proof of Correctness} In order to show that this algorithm is correct, we should show that the new RationalGRL model still satisfies all the conditions of Def.~\ref{def:grl-model}. This is clearly the case, since adding a single element does not invalidate any condition.

@@

\subsubsection{Algorithm~\ref{alg:as5} for argument scheme AS5}

\paragraph{Explanation} The procedure in Algorithm~\ref{alg:as5} takes three arguments: $g_{id}$ is the identifier of goal $G$, $T=(T_1,\ldots,T_n)$ is a list of decomposing task names, and $type\in\{and,or,xor\}$ is the decomposition type. The difficulty of this algorithm is that each of the tasks are stated in natural language, and it is not directly clear whether these tasks are already in the GRL model or not. Therefore, we have to check for each tasks whether it already exists, and if not, we have to create a new task. On line~\ref{alg:as5:1}, the set $T_{id}$ is initialized, which will contain the ids of the tasks $T_1,\ldots,T_n$ to decompose into. In the for-loop, the if-statement on line~\ref{alg:as5:2} checks whether some argument already exists for the current task $T_i$, and if so, it simply adds the identifier of the task ($t_{id}$) to the set of task identifiers $T_{id}$.\footnote{Note that the existing task may be disabled, but this does not matter, since the decomposition relation will be suppressed for disabled elements.} Otherwise (line~\ref{alg:as5:3}), a new task is created and the new identifier $id$ is added to the set of task identifiers. After the for loop on line~\ref{alg:as5:4}, an argument for the decomposition link itself is constructed, and it is added to the set of arguments $Args$.

\paragraph{Example}
Let us explain this algorithm with the XOR-decomposition of goal \texttt{Generate cars} of Figure~\ref{fig:transcripts:grl}. Suppose the following arguments are constructed already:
\begin{itemize}
\item $\{goal(0),name(0,generate\_cars)\},$
\item $\{taks(1),name(1,keep\_same\_cars\}.$
\end{itemize}
Suppose furthermore that someone wants to put forward the argument that goal \texttt{Generate cars} XOR-decomposes into tasks \texttt{Keep same cars} and \texttt{Create news cars}. Formally: $AS_5(0,\{generate\_cars,keep\_same\_cars\},xor)$. The algorithm will first set $T_{id}=\emptyset$, and then iterate over the two task names. For the first task $generate\_cars$, there does not exist an argument $\{task(t_{id}),name(t_{id},generate\_cars)\}$ yet, so a  new argument is created. Suppose the following argument is created: $\{task(2),name(2,generate\_cars)\}.$ After this, 2 is added to $T_{id}$. For the second task an argument exists already, namely $\{task(1),name(1,keep\_same\_cars)\}$, so 1 is simply added to $T_{id}$. After the for loop, we have $T_{id}=\{1,2\}$. Next, the decomposition argument is created, which is $\{decomp(3,0,\{1,2\},xor)\}$. This argument is added to $Args$ and the algorithm terminates.

@@

\paragraph{Proof of Correctness}

\subsubsection{Algorithm~\ref{alg:as6} for argument scheme AS6}

\paragraph{Explanation} The procedure in Algorithm~\ref{alg:as6} takes two arguments: $t_{id}$ is the identifier of task $T$, and $s$ is the softgoal name that is contributed to. The idea behind this algorithm is very similar to the previous one, with the difference that in the current algorithm we create a single relation, while we created a set of relations in the previous algorithm. First, the if-statement on line~\ref{alg:as6:if} checks whether the softgoal exists already, and if not, an argument is added for it. This ensures that all softgoals have corresponding arguments. After the if-statement, the argument for the contribution link is created and it is added to the set of arguments $Args$.

\paragraph{Example} Let us again illustrate this with a simple example from Figure~\ref{fig:transcripts:grl}. Suppose the following argument exists already: $\{task(0),name(0,keep\_same\_cars\}$, and suppose someone would like to add an argument that the task \texttt{Keep same cars} contributes positively to softgoal \texttt{Dynamic simulation}, i.e. $AS_6(0,dynamic\_simulation)$. The algorithm first checks whether an argument already exists for the softgoal, and when it finds out it does not exist, creates the argument $\{softgoal(1),name(1,dynamic\_simulation)\}$ and adds it to $Args$. Then, the argument for the contribution is added to $Args$ as well, which is $\{contr(2,0,1,pos)\}$.

@@

\paragraph{Proof of correctness}

@@

\subsection{Constructing GRL models}
\label{sect:rationalGRL-GRL}
\todo{F}{F,M}{It would be good to have a formal definition of how to construct models given an argumentation framework. If we define a model M at the start of this section, this should be fairly easy.}
